Forecasting interrupted time series

### 2024-05-17 fits.tex
Final changes to revised paper and response

\usepackage[section]{placeins}
\usepackage{orcidlink}
\definecolor{mypink}{RGB}{219, 48, 122}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[style=authoryear-comp,]{biblatex}
\addbibresource{references.bib}
\usepackage{bookmark}

@@

\title{Forecasting interrupted time series}
\date{20 February 2024}

@@

\nocover
  \blind
  \lfoot{\sf20 February 2024}

@@

Time series forecasting models use historical data to estimate future
values \autocite{fildes2008forecasting} based on consistent patterns
observed in the past. However, time series are sometimes disrupted by
unusual events that jeopardise the regularity of the time series
pattern. This can be defined as an \emph{interrupted time series}. Time
series data may be disrupted by a variety of reasons, ranging from
temporary policy changes to natural disasters. Outbreaks, epidemics, and
pandemics, for example, may have a considerable impact on population
behavior, causing discrepancies and interruptions of data in a variety
of industries such as healthcare, pharmaceuticals, transportation,
retail, tourism, and traffic management. Similarly, supply chain
disruptions induced by factors such as strikes or shortages cause
irregularities in sales, inventory levels, production, and delivery,
often resulting in data gaps. Transportation-related data may be
interrupted due to equipment failures, such as malfunctions or
breakdowns in vehicles, aircraft, or rail systems, resulting in missing
or inaccurate data points. Furthermore, cybersecurity breaches data
quality by possibly skewing readings in datasets relating to financial
transactions, network traffic, and user activity, complicating time
series analysis.

@@

It is important to note that in this study, we consider the problem of
forecasting interrupted time series both during and after the event.
This is a separate problem from change point detection
\autocite{aue2013structural,truong2020selective} and anomaly detection
\autocite{talagala2020anomaly}. In the situations we investigate, we
know that a change has occurred which has caused some unusual
observations, and we want to forecast what will happen next. Change
point detection refers to determining \emph{when} a change occurred
\autocite{blazquez2021review}, and anomaly detection aims to identify
unusual observations with the intention of either minimising their
value, or paying particular attention to them, and doing a root-cause
investigation. Our study focus also differs from \emph{intervention
modeling}, sometimes referred to as \emph{interrupted time series
analysis and modelling} \autocite{mcdowall2019interrupted}, which is
used when we have data about an outcome over time and the aim is to
assess the impact of an intervention, policy, or program implemented at
a specific time point. We are not trying to measure the impact of the
interruption; rather we are trying to produce sensible forecasts during
and after the interruption.

@@

In this section, we describe several possible strategies to handle
interruptions when forecasting time series data.

@@

@@

This has the advantage of retaining the memory of the past, and so the
seasonal patterns and other dynamics will be retained. This allows the
change period to be effectively modelled, provided the intervention
variables are chosen well. However, the model will assume that the
post-interruption period will be similar to the pre-interruption period,
and so the prediction intervals may be too narrow, especially if there
is a lasting effect beyond the end of the interruption.

@@

\caption{\label{fig-tourism-plot1}Short-term visitor arrivals to
Australia (monthly): Jan 2000 -- May 2023.}

@@

We apply the first four solutions discussed in the previous section to
these data, making 12 month forecasts at the end of each year from 2019
to 2022. We fit ETS, ARIMA, and dynamic ARIMA models to the data
\autocite{fpp3}, first applying a log transformation to ensure the
resulting forecasts are positive. This is possible because the
observations never reach exactly zero, with the smallest number of
visitors per month equal to 2250 in April 2020. For all forecasts, we
also show 90\% prediction intervals.

@@

\caption{\label{fig-tsol1-plot}\textbf{Highly adaptive models}.
Forecasts from ETS and ARIMA models. Neither works particularly well for
disruptions of this magnitude.}

@@

\caption{\label{fig-tsol2-plot}\textbf{Intervention model}. We use a
level shift from March 2020 to October 2022, and a ramp from October
2021 to October 2022. After October 2022, all intervention variables are
set to zero.}

@@

\caption{\label{fig-tsol3-plot}\textbf{Set to missing}. The period from
March 2020 to October 2022 is set to missing. So the first three years
of forecasts show what might have been. The fourth set of forecasts uses
observations from November and December 2022, and so the forecasts have
been adjusted downwards.}

@@

\caption{\label{fig-tsol4-plot}\textbf{Estimate what might have been}.
We replace the observations from March 2020 to October 2022 with the
average of the same month in the three years prior to March 2020.}

@@

Finally, in Figure~\ref{fig-tensemble-plot}, we show the ensemble
forecasts obtained by averaging the forecasts from the four solutions
discussed above. The first and fourth solutions carry double weight,
because both the ETS and ARIMA forecasts were included in the ensemble.
This time, no prediction intervals have been computed. The first set of
forecasts has not taken any account of the disruption, and so they are
almost identical to those obtained in the previous plots. The last set
of forecasts is reasonably good, showing what is obtained by combining
the forecasting methods used to handle the disruption period. However,
the forecasts for 2021 and 2022 are particularly poor, because they are
averaging forecasts that aim to achieve different objectives, and are
affected by the poor forecasts obtained using the dynamic regression
model.

@@

\caption{\label{fig-tensemble-plot}\textbf{Ensemble approach}, combining
the previous four approaches shown in Figures
\ref{fig-tsol1-plot}--\ref{fig-tsol4-plot}.}

@@

\FloatBarrier

@@

\caption{\label{fig-walkers}Pedestrian traffic in Melbourne, Australia,
from 2018 to 2023. The gray shaded regions indicate periods of lockdown
due to COVID-19.}

@@

\caption{\label{fig-walkers-plot1}\textbf{Highly adaptive model}: the
forecasts do not take into account the lockdown periods, but the ARIMA
model is largely able to adjust to the changing level of the series.}

@@

\caption{\label{fig-walkers-plot2}\textbf{Intervention model}: the
forecasts are based on an ARIMA model with a dummy variable indicating
when the lockdown periods occurred. Many people were in self-imposed
lockdown before the start of the first lockdown period, and were
reluctant to return to the city after it ended, resulting in poor
forecasts on either side of the first lockdown. The model was able to
adjust to the changing level of the series during subsequent lockdowns.}

@@

\caption{\label{fig-walkers-plot3}\textbf{Set to missing}: the lockdown
periods are set to missing, and the ARIMA model uses the remaining data
to produce forecasts. Thus, the forecasts are based on only previous
non-lockdown data. Naturally, the forecasts during the lockdown periods
do not reflect reality, but they can be interpreted as what might have
been.}

@@

\caption{\label{fig-walkers-plot4}\textbf{Estimate what might have
been}: the lockdown periods are replaced with estimated counts based on
an ARIMA model applied to the remaining data. Then a new ARIMA model is
fitted to the whole data set, and used to produce forecasts. As a
result, the forecasts during the lockdown periods are not true
forecasts, as they use non-lockdown data from the future.}

@@

\caption{\label{fig-walkers-ensemble-plot}\textbf{Ensemble forecasts}
based on the first two solutions shown in Figures
\ref{fig-walkers-plot1} and \ref{fig-walkers-plot2}. Other than having
some difficulty around the period of the first lockdown, the results are
relatively good.}

@@

@@

>>>>

\usepackage[section]{placeins}
\widowpenalty=10000
\clubpenalty=10000
\definecolor{mypink}{RGB}{219, 48, 122}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[style=authoryear-comp,]{biblatex}
\addbibresource{references.bib}
\usepackage{bookmark}

@@

\title{Forecasting interrupted time series}
\date{17 May 2024}

@@

\nocover
  \blind
  \lfoot{\sf17 May 2024}

@@

Time series forecasting models use historical data to estimate future
values \autocite{fildes2008forecasting} based on consistent patterns
observed in the past. However, time series are sometimes disrupted by
unusual events that jeopardise the regularity of the time series
pattern. This can be defined as an \emph{interrupted time series}. Time
series data may be disrupted by a variety of reasons, ranging from
temporary policy changes to natural disasters. Outbreaks, epidemics, and
pandemics, for example, may have a considerable impact on population
behavior, causing discrepancies and interruptions of data in a variety
of industries such as healthcare, pharmaceuticals, transportation,
retail, tourism, and traffic management. Similarly,
transportation-related data may be interrupted due to equipment
failures, such as malfunctions or breakdowns in vehicles, aircraft, or
rail systems, resulting in missing or inaccurate data points.
Furthermore, cybersecurity breaches data quality by possibly skewing
readings in datasets relating to financial transactions, network
traffic, and user activity, complicating time series analysis. Supply
chain disruptions induced by factors such as strikes or shortages cause
irregularities in sales, inventory levels, production, and delivery,
often resulting in data gaps.
\textcolor{teal}{For example, the war in Ukraine caused disruptions in sunflower oil and wheat supply chains, which were most likely driven by social media and people's recent memories of the COVID-19 toilet paper shortage. A power outage in Eastern Germany halted production at a Tesla facility and cut off electricity to a large retailer distribution centre, disrupting supplies to around 500 stores in and around Berlin for several weeks. Recent strikes in various forms of transport, including air and rail in Germany, France, and the United Kingdom, have had an impact on passenger flow and retail sales at airports and train stations. These issues highlight the significance of robust forecasting systems in addressing such disruptions effectively.}

@@

We consider the problem of forecasting interrupted time series both
during and after the event. This is a separate problem from change point
detection \autocite{aue2013structural,truong2020selective} and anomaly
detection \autocite{talagala2020anomaly}. In the situations we
investigate, we know that a change has occurred which has caused some
unusual observations, and we want to forecast what will happen next.
Change point detection refers to determining \emph{when} a change
occurred \autocite{blazquez2021review}, and anomaly detection aims to
identify unusual observations with the intention of either minimising
their value, or paying particular attention to them, and doing a
root-cause investigation. Our study focus also differs from
\emph{intervention modeling}, sometimes referred to as \emph{interrupted
time series analysis and modelling} \autocite{mcdowall2019interrupted},
which is used when we have data about an outcome over time and the aim
is to assess the impact of an intervention, policy, or program
implemented at a specific time point. We are not trying to measure the
impact of the interruption; rather we are trying to produce sensible
forecasts during and after the interruption.

@@

In this section, we describe several possible strategies to handle
interruptions when forecasting time series data.
\textcolor{teal}{The handling of time series interruptions is a complex and multifaceted issue, and the choice of the most appropriate strategy depends on the specific characteristics of the data and the nature of the interruption. Therefore, in this paper, we propose multiple strategies rather than a single approach to effectively address different situations and cases.}

@@

\textcolor{teal}{The disadvantage of smoothing and other highly adaptive techniques is that they need time to adapt after a disturbance and can eventually end up chasing noise. The latter is an intriguing issue since individuals often prefer adaptability while failing to recognise that it does not necessarily improve forecasting.}

@@

This has the advantage of retaining the memory of the past, and so the
seasonal patterns and other dynamics will be retained. This allows the
change period to be effectively modelled, provided the intervention
variables are chosen well.
\textcolor{teal}{Intervention models are useful because they enable the forecasting of similar future disruptions, but distinguishing between a disruption and a general driving factor, such as a promotion or a seasonal storm, may become subjective.}
However, the model will assume that the post-interruption period will be
similar to the pre-interruption period, and so the prediction intervals
may be too narrow, especially if there is a lasting effect beyond the
end of the interruption.

@@

\caption{\label{fig-tourism-plot1}Short-term visitor arrivals to
Australia (monthly): Jan 2000 -- May 2023. Source:
\textcite{tourismdata}. The Australian international borders closed in
March 2020 and officially reopened in February 2022.}

@@

We apply the first four solutions discussed in the previous section to
these data, making 12 month forecasts at the end of each year from 2019
to 2022. We fit ETS, ARIMA, and dynamic ARIMA models to the data
\autocite{fpp3}, first applying a log transformation to ensure the
resulting forecasts are positive. (A bias correction is also applied to
ensure the back-transformed forecasts are the means of the forecast
distributions.) A log transformation is possible because the
observations never reach exactly zero, with the smallest number of
visitors per month equal to 2250 in April 2020. For all forecasts, we
also show 90\% prediction intervals.

@@

\caption{\label{fig-tsol1-plot}\textbf{Highly adaptive models}. Each
panel shows one year of monthly forecasts from ETS and ARIMA models.
Neither works particularly well for disruptions of this magnitude.}

@@

\caption{\label{fig-tsol2-plot}\textbf{Intervention model}. Each panel
shows one year of monthly forecasts from an intervention model
containing a level shift from March 2020 to October 2022, and a ramp
from October 2021 to October 2022. After October 2022, all intervention
variables are set to zero.}

@@

\caption{\label{fig-tsol3-plot}\textbf{Set to missing}. Each panel shows
one year of monthly forecasts from an ARIMA model. The period from March
2020 to October 2022 is set to missing. So the first three years of
forecasts show what might have been. The fourth set of forecasts uses
observations from November and December 2022, and so the forecasts have
been adjusted downwards.}

@@

\caption{\label{fig-tsol4-plot}\textbf{Estimate what might have been}.
Each panel shows one year of monthly forecasts from ETS and ARIMA
models. We have replaced the observations from March 2020 to October
2022 with the average of the same month in the three years prior to
March 2020.}

@@

@@

\caption{\label{fig-tensemble-plot}\textbf{Ensemble approach}, combining
the previous four approaches shown in Figures
\ref{fig-tsol1-plot}--\ref{fig-tsol4-plot}. The point forecasts are the
average of the four approaches, while the prediction intervals are
formed by combining the four forecast distributions at each horizon to
form a mixture distribution.}

@@

Finally, in Figure~\ref{fig-tensemble-plot}, we show the ensemble
forecasts obtained by averaging the forecasts from the four solutions
discussed above. The first and fourth solutions carry double weight,
because both the ETS and ARIMA forecasts were included in the ensemble.
The prediction intervals are formed from the mixture distributions
obtained by combining the four forecast distributions for each time
horizon. The first set of forecasts has not taken any account of the
disruption, and so they are almost identical to those obtained in the
previous plots. The last set of forecasts is reasonably good, showing
what is obtained by combining the forecasting methods used to handle the
disruption period. However, the forecasts for 2021 and 2022 are
particularly poor, because they are averaging forecasts that aim to
achieve different objectives, and are affected by the poor forecasts
obtained using the dynamic regression model.

@@

\caption{\label{fig-walkers}Average pedestrian traffic in Melbourne,
Australia, from 2018 to 2023, based on automated hourly counts from
sensors around the city. The counts are averaged over
\texttt{r\ ngoodsensors} sensors, each of which had no more than 30 days
of missing data. The gray shaded regions indicate periods of lockdown
due to COVID-19. Source: \textcite{pedestrians}}

@@

\caption{\label{fig-walkers-plot1}\textbf{Highly adaptive model}:
Rolling 7-day forecasts of daily pedestrian data, made each week, using
ARIMA models. The forecasts do not take into account the lockdown
periods, but the ARIMA models are largely able to adjust to the changing
level of the series.}

@@

\caption{\label{fig-walkers-plot2}\textbf{Intervention model}: Rolling
7-day forecasts of daily pedestrian data, made each week, using ARIMA
models with a dummy variable indicating when the lockdown periods
occurred. Many people were in self-imposed lockdown before the start of
the first lockdown period, and were reluctant to return to the city
after it ended, resulting in poor forecasts on either side of the first
lockdown. The model was able to adjust to the changing level of the
series during subsequent lockdowns.}

@@

\caption{\label{fig-walkers-plot3}\textbf{Set to missing}: Rolling 7-day
forecasts of daily pedestrian data, made each week, using ARIMA models,
where the lockdown periods are set to missing. The ARIMA models use the
remaining data to produce forecasts. Thus, the forecasts are based on
only previous non-lockdown data. Naturally, the forecasts during the
lockdown periods do not reflect reality, but they can be interpreted as
what might have been.}

@@

\caption{\label{fig-walkers-plot4}\textbf{Estimate what might have
been}: Rolling 7-day forecasts of daily pedestrian data, made each week.
The lockdown periods are replaced with estimated counts based on an
ARIMA model applied to the remaining data. Then a new ARIMA model is
fitted to each data set, and used to produce forecasts. As a result, the
forecasts during the lockdown periods are not true forecasts, as they
use non-lockdown data from the future.}

@@

\caption{\label{fig-walkers-ensemble-plot}\textbf{Ensemble forecasts}
combining the first two solutions shown in Figures
\ref{fig-walkers-plot1} and \ref{fig-walkers-plot2}. Other than having
some difficulty around the period of the first lockdown, the results are
relatively good.}

@@

\textcolor{teal}{When forecasting interrupted time series data, exogenous variables associated with specific events should be identified and incorporated where possible, regardless of which approach is employed, since they may improve forecasting model performance.}

@@

\section*{Acknowledgment}\label{acknowledgment}
\addcontentsline{toc}{section}{Acknowledgment}

We thank Dr Stephan Kolassa and several referees for their thoughts and
feedback on the first version of our paper. These comments have helped
us to refine and improve our work.
